Ion pairing controls rheological properties of ``processionary'' polyelectrolyte hydrogels

### 2016-02-15 ion_pairing_paper.tex
harder -> well hydrated. Reorganize the pi-pi vs pi+-pi+ sentence.

9\,cm by 3.5\,cm

@@

@@

In the present report, we investigate the role of the ion pair dissociation equilibria rate of the poly(cationic) body on the mechanical properties of the hydrogel. In a first part we describe how we take advantage of our postfunctionalization approach to play systematically with the nature (hardness/softness) of the cationic repeating unit and the ion-paring strength of the associated counter-ion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}). In a second part, we show the qualitative and quantitative variations of mechanical properties of the aqueous dispersions obtained from these well-characterized polymers. In a third part, we develop a microscopic model based on the idea that at low dissociation rates, cross-links (defined as a polymer that is connected to three or more others) are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular chain. We call this behaviour ``processionary'' in analogy to pine processionary caterpillar (\textit{Thaumetopoea pityocampa}) behaviour\cite{Fabre1916}.%Finally, we discuss our results and possible applications of such smart hydrogels.

@@

\subsection{Water swelling, ion-pairing and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer composition. With bromide or iodide conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With fluoride counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 8\%wt suspension as a biphasic mixture of its individual components.

The above dependence of gel formability on the nature of the counterions could be rationalized on the basis of ion-pairing efficiency. According to the Hard and Soft Acid and Bases (HSAB) theory, stronger association is expected between a ``softer'' cationic moiety and a ``harder'' halide counterions\cite{Goossens2009}. Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties: whereas pyrrolidinium features a ``hard'' positive charge very localized on the quaternary nitrogen, imidazolium has a much more delocalized ``softer'' charge. Therefore charge dissociation in water and thus solubility of the polymer should be easier with larger halide and pyrrolidinium. As a consequence, while dissolution is observed for the four polymers with bromide and iodide counterions and for \ce{PPyr+Cl-}, in which ion dissociation takes place to a sufficient extent, neither \ce{PIm+Cl-} nor the two polymers with fluoride counterions, for which ion pairs are expected to be strongly bound, can be dissolved.

Such an insolubility in absence of charge dissociation indicates that the chain is in poor solvent. Locally, the monomers are condensed in collapsed globules to minimize the contacts with solvent, and solubility on larger scales can only be achieved if the polymer bears enough charges\cite{Khokhlov1980,Raphael1990}. To estimate the $\Theta$-temperature, that delimit bad and good solvent behaviour, we held the two fluoride polymers in water at boiling temperature overnight. No dissolution or swelling was observed, implying that $\Theta>\SI{100}{\celsius}$. %Nonetheless, due to charge dissociation, five of our polyelectrolytes are soluble at \SI{25}{\celsius}.

@@

\begin{acknowledgement}
The authors are grateful for the financial support of Université de Lyon through the program ``Investissements d'Avenir'' (ANR-1 1-IDEX-0007). S.M. and N.T. acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007–2013)/ERC grant agreement No.258803.
\end{acknowledgement}

>>>>

%9\,cm by 3.5\,cm
\includegraphics{TOC2.png}

@@

Contrary to a small electrolyte that exhibit only dissociated ions, polyelectrolytes in solution are in general not fully dissociated, with a substantial fraction of the counterions bound to the polymer\cite{Manning1979}. The resulting net charge of the chain governs the physical properties of the polyelectrolytes solution\cite{Rubinstein1996}, first of all the ability of the polymer to dissolve. counterion condensation is the physical bounding or adsorption of counterions near the polymer chain. Factors influencing this process have been known since the 1880's but are still being rationalized. \citet{Hofmeister1888} was the first to propose a systematic ranking of ions, based on their propensity to promote the coalescence of egg white. This so-called Hofmeister series has since proved quite universally valid\cite{Zhang2010a}, including multi-charged polymers such as naturally occurring proteins or synthetic polyelectrolytes as well as charged colloidal particles\cite{Schwierz2010,Oncsik2015} or soft matter interfaces in general\cite{Leontidis2014}. However, depending on the nature of the polymer (hydrophilic or hydrophobic, anionic or cationic), the Hofmeister series can be direct (in short, well hydrated ions promote condensation) or reverse\cite{Schwierz2010,Oncsik2015}.

@@

In the present report, we investigate the role of the ion pair dissociation equilibria rate of the poly(cationic) body on the mechanical properties of the hydrogel. In a first part we describe how we take advantage of our postfunctionalization approach to play systematically with the nature (aromatic or not) of the cationic repeating unit and the associated counterion (\ce{F-}/\ce{Cl-}/\ce{Br-}/\ce{I-}) varied along the Hofmeister series. In the second part and third part, we show respectively the qualitative change in gel formability and the quantitative variations of mechanical properties of the aqueous dispersions obtained from these well-characterized polymers. In a fourth part, we develop a microscopic model based on the idea that at low dissociation rates, cross-links (defined as a polymer that is connected to three or more others) are not separated by a single macromolecule but by several (up to hundreds) polymers in a supramolecular chain. We call this behaviour ``processionary'' in analogy to pine processionary caterpillar (\textit{Thaumetopoea pityocampa}) behaviour\cite{Fabre1916}.%Finally, we discuss our results and possible applications of such smart hydrogels.

@@

\subsection{Water swelling, counterion condensation and solvent quality}
Water-swelling properties turn out to be extremely dependent on the polymer composition. With bromide or iodide conterions, both poly(pyrrolidinium) and poly(immidazolium) polymers afford homogeneous and optically transparent gels upon swelling with deionised water. We observe a limited swelling for \ce{PPyr+Cl-}, resulting in a granular, inhomogeneous gel. With fluoride counterions, or in the case of \ce{PIm+Cl-}, we observe neither swelling nor dissolution. We recover a 8\%wt suspension as a biphasic mixture of its individual components.

The above dependence of solubility on the nature of the counterions follows the direct Hofmeister series. The strong condensation of \ce{F-} on the chain decreases the effective positive charge of the polymer and thus reduces solubility. Conversely, larger, poorly hydrated ions like \ce{I-} are less adsorbed, resulting in more dissociated charges and higher solubility. As a consequence, while dissolution is observed for the four polymers with bromide and iodide counterions, in which ion dissociation takes place to a sufficient extent, the two polymers with fluoride counterions, for which couterion condensation is expected to be strong, cannot be dissolved.

However Hofmeister series alone cannot explain the difference between soluble \ce{PPyr+Cl-} and insoluble \ce{PIm+Cl-}. We have to suppose that counterions are more strongly condensed near imidazolium than near pyrrolidinium. Indeed, although both cationic in nature, pyrrolidinium and imidazolium ions have markedly different properties. Because of their aromatic ring, imidazolium are particularly prone to promote a variety of supramolecular interactions which strongly contribute to their physicochemical properties, such as \ce{\pi+-\pi} or anion$-\pi^+$.  More recently, \ce{\pi+-\pi+} interactions have also been identified as a distinctive driving force for imidazolium dimerization\cite{Geronimo2011}. Because of strong intercoulombic repulsion, \ce{\pi+-\pi+} are primarily weaker than their \ce{\pi-\pi} counterparts. However it has been established that, in the presence of couterion, the stabilizing effect of \ce{\pi+-\pi+} interaction could reach magnitudes largely exceeding that of the latter\cite{Geronimo2011}. In particular, dimerization of imidazolium-chloride ion pairs have been the object of recent studies, and it has been shown that the involvement of the negatively charged counterion was essential in maximizing the stabilization of the \ce{\pi+-\pi+} complex by minimizing coulombic repulsion between the imidazolium moieties\cite{Matthews2014,Gao2015a}. It is likewise well documented that covalent incorporation of interacting group within a polymer chain strongly favours intramolecular interactions between these groups, by increasing their so-called effective molarity\cite{Li2003,Mulder2004,Huerta2013}. It is therefore not surprising that both effects participate in making the imidazolium-anion interaction stronger than pyrrolidinium at the detriment of solubility.

Such an insolubility in absence of charge dissociation indicates that the chain is in poor solvent. Locally, the monomers are condensed in collapsed globules to minimize the contacts with water molecules, and solubility on larger scales can only be achieved if the polymer bears enough charges\cite{Khokhlov1980,Raphael1990}. To estimate the $\Theta$-temperature, that delimit bad and good solvent behaviour, we held the two fluoride polymers in water at boiling temperature overnight. No dissolution or swelling was observed, implying that $\Theta>\SI{100}{\celsius}$. %Nonetheless, due to charge dissociation, five of our polyelectrolytes are soluble at \SI{25}{\celsius}.

@@

\begin{acknowledgement}
The authors are grateful for the financial support of Université de Lyon through the program ``Investissements d'Avenir'' (ANR-1 1-IDEX-0007). S.M. and N.T. acknowledge funding from the European Research Council under the European Union's Seventh Framework Program (FP7/2007–2013)/ERC grant agreement No.258803. The Graphical TOC entry is derived from a picture by Arturo Reina, CC BY-SA 3.0, \url{https://commons.wikimedia.org/w/index.php?curid=282898}.
\end{acknowledgement}
